Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-09-27 papers/paper1/abstract.tex
General revision and added conclusion.

\begin{abstract}
The ability to realize sophisticated distributed systems such as those found in cyber-physical, pervasive, or enterprise computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly and easily in reusable software.
% By ``different forms'' we mean single-threaded, multi-threaded, time triggered, event triggered, etc.

To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
We develop a component model for asynchronous and concurrent systems based on the I/O automata formal model and dynamic composition.
We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.

@@

We evaluate the ability to execute actions concurrently in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
\end{abstract}

>>>>

\begin{abstract}
The ability to realize sophisticated distributed systems such as those found in cyber-physical, pervasive, or enterprise computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly in reusable software.
% By ``different forms'' we mean single-threaded, multi-threaded, time triggered, event triggered, etc.

To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
We develop a component model for asynchronous and concurrent systems based on I/O automata and dynamic composition.
We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.

@@

We evaluate concurrency in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
\end{abstract}

### 2011-09-27 papers/paper1/case_study.tex
General revision and added conclusion.

\paragraph*{Translating the asynchronous LCR automaton into ioa++}
The first version of the code for the asynchronous LCR automaton was translated directly from a listing in~\cite{lynch1996distributed}.
The final version contains an additional action to reinitialize the automaton and modifications that allow an automaton that elected itself leader to nullify its election.
The motivation for these modifications will be explained when we discuss its use in a real system.
The listing in~\cite{lynch1996distributed} consists of approximately 24 lines in a formal language that must be translated to a production language.
The raw listing for the final version consists of 121 lines (5x).
Scheduling accounts for 22 of the 121 lines (20\%) and is the main source of code that cannot be traced back to the formal listing.
On average, every local action contributed 6 lines of scheduling code and every input action contributed 3 lines of scheduling code.

@@

\paragraph*{Unit testing with simulation}
After the asynchronous LCR automaton was implemented, we developed a simulation to test it.
The examples/asynch\_lcr.cpp file contains the driver for the simulation.
The simulation creates a unidirectional ring of a specified size consisting of alternating asynch\_lcr\_automaton and channel\_automaton.
(The channel\_automaton was also a direct translation from~\cite{lynch1996distributed}.)

>>>>

\paragraph*{Translating the asynchronous LCR automaton into ioa++}
The first version of the code for the asynchronous LCR automaton was translated directly from a listing in~\cite{lynch1996distributed}.
The final version contains an additional action to reinitialize the automaton and modifications that allow an automaton that elected itself leader to nullify its election.
The motivation for these modifications will be explained when we discuss its use in a real system.
The listing in~\cite{lynch1996distributed} consists of approximately 24 lines in a formal language that must be translated to a production language.
The raw listing for the final version consists of 121 lines (\~5x).
Scheduling accounts for 22 of the 121 lines (20\%) and is the main source of code that cannot be traced back to the formal listing.
On average, every local action contributed 6 lines of scheduling code and every input action contributed 3 lines of scheduling code.

@@

\paragraph*{Unit testing with simulation}
After the asynchronous LCR automaton was implemented, we developed a simulation to test it.
The examples/asynch\_lcr.cpp file contains the driver for the simulation.
The simulation creates a unidirectional ring of a specified size consisting of alternating asynch\_lcr\_automaton and channel\_automaton.
The channel\_automaton was also a direct translation from~\cite{lynch1996distributed}.

### 2011-09-27 papers/paper1/component_model.tex
General revision and added conclusion.

\subsection{I/O Automata\label{io_automaton}}

In this subsection we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected for execution.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.

Together, an automaton's output and input actions constitute its \emph{external signature}~\cite{lynch1996distributed}.
An automaton communicates with other automata through the actions in its external signature.
I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effects of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
Composing automata that have identically named output actions would be ambiguous and thus is not allowed.
The result of composition is a new automaton, so that the principles for reasoning about the behavior of a composed system are the same as the principles for reasoning about a single automaton.
Often, the properties of the resulting aggregate automaton can be reasoned about using only the properties of the constituent automata.
%% A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.

@@

@@

The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

System actions can fail for a variety of reasons.
A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{io_automaton} would be violated.
An unbind fails if the binding does not exist.
Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.

@@

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the set defined in Section~\ref{io_automata} \emph{and} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{io_automata} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

>>>>

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
%% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
We then describe extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time.

\subsection{Background: I/O Automata}

The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\ifjournal
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
\fi
An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
\ifjournal
Often, a local action consists of a precondition and an effect.
The precondition is a predicate over the state variables that enables the effect when selected for execution.
The effect is a function that computes the next state of the automaton from the current state.
An input action presented in this style consists solely of an effect.
\fi
An automaton's output and input actions constitute its \emph{external signature}~\cite{lynch1996distributed}.

An automaton communicates with other automata through the actions in its external signature.
I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effects of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
Composing automata that have identically named output actions would be ambiguous and thus is not allowed.
The result of composition is a new automaton, so that the principles for reasoning about the behavior of a composed system are the same as the principles for reasoning about a single automaton.
Often, the properties of the resulting aggregate automaton can be reasoned about using only the properties of the constituent automata.
%% A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.

@@

\subsection{Component Structure\label{component_structure}}

@@

The event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

@@

System actions can fail for a variety of reasons.
A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{component_structure} would be violated.
An unbind fails if the binding does not exist.
Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.

@@

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the set defined in Section~\ref{component_structure} \emph{and} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

### 2011-09-27 papers/paper1/conclusion.tex
General revision and added conclusion.

\begin{itemize}
  \item We are going to use it to build the substrate.
  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
  \item We can take advantage of multi-core in a very straight-forward way
  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
\end{itemize}

>>>>

I/O automata are a good foundation for asynchronous and concurrent components due to independent state, well-defined interfaces, and well-defined interactions under composition.
The ioa++ framework facilitates concurrent execution via dynamic composition and the degree of concurrency is limited by the interactions of the automata and the overhead of the framework.
The two key challenges when moving from the formal model to an actual implementation were introducing features for managing dynamic constellations of automata and providing a concrete scheduling mechanism.
We believe that attempts to implement other formal models, e.g., UNITY, will have similar experiences and thus encounter the differences between theory and practice.
We agree with~\cite{georgiou2009automated} that having a representation of a model that can be compiled is beneficial because it allows one to reason about the program directly from the source code and helps to close the gap between mental model and code listing.

Our primary goal moving forward is to use the ioa++ framework to gain experience building real systems with I/O automata.
The ioa++ has no support for static composition.
We hope to resolve this weakness in the future.
A significant open problem in ioa++ is the design of an efficient dispatcher and scheduler(s).

%% \begin{itemize}
%%   \item We are going to use it to build the substrate.
%%   \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
%%   \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
%%   \item We can take advantage of multi-core in a very straight-forward way
%%   \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
%%   \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
%%   \item Static composition
%%   \item Grandparents
%%   \item We own the event loop
%%   \item The procedure
%% \end{itemize}

### 2011-09-27 papers/paper1/design.tex
General revision and added conclusion.

ioa++ is an implementation in C++ of the component model described in Section~\ref{component_model} for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

@@

Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.

@@

\paragraph*{File descriptor events}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

>>>>

ioa++ is an implementation of the component model described in Section~\ref{component_model} for POSIX environments in C++.
Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

@@

Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple-reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.

@@

\paragraph*{File descriptor events}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

### 2011-09-27 papers/paper1/evaluation.tex
General revision and added conclusion.

ioa++ permits independent actions to be executed concurrently.
The goal of this evaluation is (1) to show the concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.

@@

Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
Confidence intervals are shown with largest being 0.002155 for Figure~\ref{overhead1} and 0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes to 1 as the system transitions from acquiring two locks to one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.

>>>>

ioa++ permits independent actions to be executed concurrently.
The goal of this evaluation is (1) to show that concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.

@@

Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
Confidence intervals are shown with largest being 0.002155 for Figure~\ref{overhead1} and 0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes to 1 and the system transitions from acquiring two locks to one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.

### 2011-09-27 papers/paper1/programming_model.tex
General revision and added conclusion.

\paragraph*{Declaring an automaton}
All automata inherit from the common base class ioa::automaton which implements the system action interface.
The following code declares the asynch\_lcr\_automaton:
\ifjournal
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
\fi
The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

\paragraph*{Declaring and initializing state variables}
The state variables of an automaton are expressed as member variables of the corresponding class.
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
\ifjournal
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Notice that the member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
State variables are initialized using a constructor.
For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue:
\ifjournal
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
\fi
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
The schedule member function is discussed in the paragraph on scheduling.

\paragraph*{Declaring/defining actions}
Actions are member variables that model the action concept.
Actions contain methods for preconditions, effects, and scheduling and traits that indicate the actions's type (input, output, internal), value status, value type, parameter status, and parameter type.
The traits are used to guide the execution of the methods by the model and supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
For example, the send action of the asynch\_lcr\_automaton is declared:
\ifjournal
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly save the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata belong in a public section.

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
For the asynch\_lcr\_automaton, the scheduling function is:
\ifjournal
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
\fi
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.

@@

\paragraph*{Dynamic composition}
The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\ifjournal
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\ifjournal
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

>>>>

\paragraph*{Declaring an automaton}
All automata inherit from the common base class ioa::automaton which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
\fi
The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

\paragraph*{Declaring and initializing state variables}
The state variables of an automaton are expressed as member variables of the corresponding class.
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
\ifjournal
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
State variables are initialized using a constructor.
For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue.
\ifjournal
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
\fi
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
The schedule member function is discussed in the paragraph on scheduling.

\paragraph*{Declaring/defining actions}
Actions are member variables that model the action concept.
Actions contain methods for preconditions, effects, and scheduling and traits that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
The traits are used to guide the execution of the methods by the model and supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

A number of wrappers and macros are available for declaring actions.
%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly save the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata belong in a public section.

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
\ifjournal
For the asynch\_lcr\_automaton, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
\fi
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.

@@

\paragraph*{Dynamic composition}
The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
\fi

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi
